### 2014-11-06 D1-3.tex
Added link between Recommendations and Implementations.

While the collection of this data is necessary for providing the
advanced services that Live+Gov aims to deliver, at he same time, the
available raw data can be used to draw a very detailed picture of the
private life of the citizen. For instance can GPS location tracking be
used to reveal shopping habits (e.g. when a car seller is visited) and
associations to political groups (when a meeting is
attended). Accelerometer data can be used to infer medical conditions
like walking disabilities.

Images can contain faces of nearby persons to with whom the citizen is
associated. All this data is highly sensitive to the citizens privacy
and can be used against the citizen if it falls in the wrong hands.

The great importance of protecting the citizens privacy should be
apparent from these examples. The European Union, as well as many
other countries in the past, has set out a number of directives that
regulate the collection, processing and use of privacy sensitive
data. The most important legislation on the European Level is the Data
Protection Directive 95/46/EC [1], which is also summarized in the
European Charta of Fundamental Rights [2] of EU citizens:

@@

>>>>

While the collection of this data is necessary for providing the
advanced services that Live+Gov aims to deliver, at he same time, the
available raw data can be used to draw a very detailed picture of the
private life of the citizen. For instance can GPS location tracking be
used to reveal shopping habits (e.g. when a car seller is visited) and
associations to political groups (when a meeting is
attended). Accelerometer data can be used to infer medical conditions
like walking disabilities. Images can contain faces of nearby persons
to with whom the citizen is associated. All this data is highly
sensitive to the citizens privacy and can be used against the citizen
if it falls in the wrong hands.

The great importance of protecting the citizens privacy should be
apparent from these examples. The European Union, as well as many
other countries in the past, has set out a number of directives that
regulate the collection, processing and use of privacy sensitive
data. The most important legislation on the European Level is the Data
Protection Directive 95/46/EC (cf. Section \ref{sec:EUDIR}), which is
also summarized in the European Charta of Fundamental Rights
(cf. Section \ref{sec:FR}) of EU citizens:

@@

* Discussed Legal and ethical aspects of privacy
* Made Privacy Security Analysis of Live+Gov System
* Presented several Privacy Protection Measures that have been implemented.
* Most importantly the Privacy Dashboard, which gives the citizen control over their data.

### 2014-11-06 Implementations.tex
Added link between Recommendations and Implementations.

\chapter{Implementation of Privacy Protection Measures}
\section{Privacy Policy} \label{sec:PP}

In complience legal requirements of the Data Protection Directive (cf. Section \ref{EUDIR}) we include the following
Privacy Policy on our Mobile Application and Website.

@@

\section{Privacy Dashboard}
The Live+Gov Privacy Dashboard is a software component that allows the
citizens to view and control the data about themselves that is stored
in the Sensor Data Storage server.  The foundation of the Privacy
Dashboard component is the inspection tool that was introduced in D1.1
as a means to perform basic quality controlling of the collected data.

\pagebreak

\section{Transfer Encryption (HTTPS/SSL)}
The data transfer from the mobile device to the Data Storage component
as well as the transfer to the Privacy Dashboard relay on HTTP. In
order to avoid eavesdropping by externals that have access to the
communication channels, we secured all HTTP connections with SSL
encryption.

@@

@@

\section{Anonymization of trips}
The Sensor Data Storage Service follows a privacy by design guideline
in that it does not store personal information like names and email
address at all.

@@

>>>>

\chapter{Implementation of Privacy Protection Measures}
\label{chap:impl}

In this chapter we explain the measures we have taken in order to
protect the citizens privacy. We have used the recommendations derived
from the Security Analysis performed in Chapter \ref{chap:analysis} as
a guideline.

\section{Privacy Dashboard}
The Live+Gov Privacy Dashboard is a software component that allows the
citizens to view and control the data about themselves that is stored
in the Sensor Data Storage server. It implements privacy
recommendation R1 of Section \ref{sec:prec}.  The foundation of the
Privacy Dashboard component is the inspection tool that was introduced
in D1.1 as a means to perform basic quality controlling of the
collected data.

\section{Privacy Policy} \label{sec:PP}
Implementing Privacy Recommendation R1 of Section \ref{sec:prec} and
following legal requirements of the Data Protection Directive
(cf. Section \ref{EUDIR}) we include the following Privacy Policy on
our Mobile Application and Website.

@@

\section{Transfer Encryption (HTTPS/SSL)}
The data transfer from the mobile device to the Data Storage component
as well as the transfer to the Privacy Dashboard relay on HTTP. In
order to avoid eavesdropping by externals that have access to the
communication channels, we secured all HTTP connections with SSL
encryption. This measure implements the security recommendation R7
Securing Communiction Channels of Section \ref{sec:prec}.

@@

\section{Security of Infrastructure}
The servers hosting the Sensor Data Storage component have been
secured using the following security measures:
\begin{itemize}
  \item Firewall. A Firewall restricts access to the minimal necessary
    ports allowing upload and inspection of the data as well, database
    access by our consortium partners and maintenance over SSH.
  \item Security Upgrades.  The operating system (Ubuntu Linux 12.04)
    is configured to install security updates of all installed
    packages on a daily basis.
  \item Access Rules. Access to the System (via SSH) and the database
    is restricted to a minimal amount of persons. Those persons have
    been advised to use strong passwords.
\end{itemize}
These measures are a basic implementation of recommendation R6.

@@

\section{Pseudonymization of trips.}
The Sensor Data Storage Service follows a privacy by design guideline
in that it does not store personal information like names and email
address at all, only pseudonyms in the form of user-ids.

@@

This measure implements part of the security recommendation R5
Anonymize personal data of Section \ref{sec:prec}.

### 2014-11-06 PrivacyProtection-2014-09-09.tex
Added link between Recommendations and Implementations.

\subsection{EU Charter of Fundamental Rights}
The Treaty of Lisbon of 2007, which introduced the fundamental functioning of the European Union
includes the \emph{Charter of Fundamental Rights of the Europen Union} \cite{EUFR2010}.
This Charter summarizes the full range of civil, political and economic rights of EU citizens and contains the following two articles that safeguard the privacy of the citizen.

@@

\chapter{Privacy Analysis of Live+Gov Systems}

@@

%TODO: Update Threat Table
\input{./figures/Threats}

%0. Insufficient Control Features
%-------------------
%Description:    The Service Provider does not offer tools for the citizen to control his data.
%COI:            Privacy vs. System Complexity
%Vulneratbility: missing privacy wareness
%Asset:          All collected data
%
%Likeliness: Always
%Loss: Low (Control is lost, no direct harm done)
%Risk:      5
%
\textbf{Insufficient Control Features.}
The System Provider does not offer tools for the Citizen to control his data.
However, besides the missing features he provides a safe and secure system.
This threat contradicts the definition of privacy as control over one's information about oneself.
As soon as collected data of Citizens is stored on System Provider servers, all control over that data is lost.
This is not necessarily due to bad intention, System Providers may simply have forgotten to include such features during the development process.
Although, control capabilities for Citizens add to the system complexity, which could motivate to omit those. 
Therefore interests of Citizens and System Providers are in conflict, namely it is the Citizen's \textit{Privacy vs. System Complexity} for System Providers.
This is an abstract threat provoked by a general \textit{Missing Privacy Awareness} in the minds of all actors in the Live+Gov context.

%1. Excessive Data Mining
%------------------------
%Description: The Service Provider and/or the Local Authority extract more private information from the collected data, than the citizen agreed to.
%E.g. the Service Provider wants to test a new product, and uses old data collected for a different purpose.
%E.g. the Local Authority wants to analyze collected data from fare evasions.
%
%COI: Privacy vs. Buisness Intelligence of Local Authority, Privacy vs. Financial Profit of Serivce Provider.
%Vuln.: Lax data handling policies; Unaware citizen; Weak law enforcement
%Affected Data: Potential Information extracted from raw data.
%
%Likeliness: High
%Loss:      Medium (Control is lost, no direct harm done)
%Risk:      5
%
\textbf{Excessive Data Mining.}
The System Provider and/or the Local Authority secretly extract more private information from the collected data, than the Citizen agreed to.
But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties.
This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection.
Or for a Local Authority, who wants to analyze the data collection regarding fare evasion.
However, the Citizen has not agreed to such data processing nor could he, since it is conducted secretly.
This disables a Citizen to control his data adequately.
Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Local Authority}
The threat can be provoked by either lax data handling policies of both System Providers and Local Authorities, or a weak law enforcement of existing supervision.
But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.

%2. Data Theft
%-------------
%Description: External infiltrates infrastructure in order to steal personal data and sell it on the black market. Also the external might be motivated politically and wants to harm the reputation of the Service Provider or the Local Authority.
%
%COI:         Privacy vs. Financial Profit
%             Reputation of Service Provider vs. Political Activism
%             Reputation of Local Authority vs. Political Activism
%Vuln.:       Insecure Infrastructure, Insecure Communication
%Affected Data:  Data transmitted or stored in data center: GPS trips, HAR-records.
%
%Likeliness: Low-Medium
%Loss: High
%Risk: 4-5
%
\textbf{Data Theft}
An External infiltrates infrastructure in order to steal personal data and sell it on the black market.
Also the External might be motivated politically and wants to harm the reputation of the System Provider or the Local Authority.
Anyways, this threat would be manifested through a technical attack on either hardware (Packet Capture) or software (buffer overflow, SQL injection).
Such a successful attack could harm the reputation of both System Provider and Local Authority.
The technical competence of System Providers would be proven wrong, therefore they would lose professional reputation.
Local Authorities would lose their political reputation, as it would seem like they had endangered data of Citizens, which lose complete control.
Thus this threat is defined by three conflicts: \textit{Privacy vs. Financial Profit}, \textit{ Reputation of Service Provider vs. Political Activism} and \textit{Reputation of Local Authority vs. Political Activism}.
Also this threat describes the classical scenario, where attacks are provoked by \textit{Insecure Infrastructure} (SQL injection) and \textit{Insecure Communication} (Packet Capture).

%3. Surveillance
%---------------
%COI:           Privacy vs. External Power
%Vulnerability: Insecure Infrastructure, Insecure Communication
%Affected Data:         Data transferred and stored in data center.
%
%Likeliness: Medium (Government*), Low (Third parties)
%Loss: High (Control is lost, potential harm by due to Government action)
%Risk:      6/4
%
%* Government, has also other tools at it's disposal
%
%
\textbf{Surveillance} External infiltrates infrastructure in order to
obtain direct about the citizen. In this scenario the external is
supposed to have some direct relationship to the citizen which
motivates his interest to obtain personal information. Examples could
be a public institution that wants to gain information about planned
activities of the citizens (e.g. Nixon's Watergate scandal or the
recent prosecution of Guardian journalists by GHCQ). Another example
is an insurance company that seeks to get information about the
citizens life-style in relation to the insured risk, like car
accidents or health hazards.

@@

%4. Information Leak
%-------------------
%COI:         Privacy vs. Financial Profit
%             Reputation of Service Provider vs. Political Activism
%             Reputation of Local Authority vs. Political Activism
%Vuln.:       Unhappy Employee, Lax Access Rules.
%Affected Data:       Data stored in data center.
%
%Likeliness: Low
%Loss: High (public leak)
%Risk: 4
%

\textbf{Information Leak}
Like an external person in Scenario (2) an employee of the service provider has selfish interests to gain money, make political statements or harm his employer.
In order to pursue this interest he can steal personal data and sell it or release it to the public.

%5. Social Engineering
%---------------------
%Description: 
%COI:         Privacy vs. Financial Profit
%             Reputation of Service Provider vs. Political Activism
%             Reputation of Local Authority vs. Political Activism
%Vuln.:       Unhappy Employee, Lax Access Rules.
%Affected Data:       Data stored in data center.
%
%Likeliness: Low
%Loss: High
%Risk: 4
%
\textbf{Social Engineering}
An External can manipulate Local Authorities or Service Providers to
give them personal information that he can then sell or disclose.

@@

For the quantification of the loss in case of a threat scenario we use the following rough calibration:
\begin{itemize}
\item 3: High. Leak of information to Peers which impacts citizen.
  E.g. spouse gets GPS data and finds out cheating.
  Subjects is added to No-Fly list due to Government spying.
  Public leak of private information.
\item 2: Medium. Undisclosed processing of personal data or disclosure to third parties that are unrelated to subject.
\item 1: Low. Loss of control over data.
\item 0: None
\end{itemize}

For the quantification of likeliness the following scale is used:
\begin{itemize}
\item 4: always the case.
\item 3: maybe occure once in 10
\item 2: maybe occure once in 100
\item 1: maybe occure once in 1 million cases.
\item 0 (impossible)
\end{itemize}

@@

%TODO:
Insert Table with Likeliness / Loss / Risk Assesment.

\subsubsection{Security Recommendations and Requirements}
\textbf{Legal Requirements}

Privacy Policy

Processing Journal
\begin{itemize}
\item name of the controller
\item purpose of processing
\item description of the data categories
\item recipients of the data if disclosed
\item transfer to third countries
\item description of security of processing
\end{itemize}
The journal is available for anybody to access via Email.
Reporting to Data Protection Authority (Bundesamt für Datenschutz).

>>>>

\subsection{EU Charter of Fundamental Rights}\label{sec:FR}
The Treaty of Lisbon of 2007, which introduced the fundamental functioning of the European Union
includes the \emph{Charter of Fundamental Rights of the Europen Union} \cite{EUFR2010}.
This Charter summarizes the full range of civil, political and economic rights of EU citizens and contains the following two articles that safeguard the privacy of the citizen.

@@

\chapter{Privacy Analysis of Live+Gov Systems}
\label{chap:analysis}

@@

% \label{figure:Threats}
\input{./figures/Threats}

\textbf{T1. Insufficient Control Features.}
The System Provider does not offer tools for the Citizen to control his data.
However, besides the missing features he provides a safe and secure system.
This threat contradicts the definition of privacy as control over one's information about oneself.
As soon as collected data of Citizens is stored on System Provider servers, all control over that data is lost.
This is not necessarily due to bad intention, System Providers may simply have forgotten to include such features during the development process.
Although, control capabilities for Citizens add to the system complexity, which could motivate to omit those. 
Therefore interests of Citizens and System Providers are in conflict, namely it is the Citizen's \textit{Privacy vs. System Complexity} for System Providers.
This is an abstract threat provoked by a general \textit{Missing Privacy Awareness} in the minds of all actors in the Live+Gov context.

\textbf{T2. Excessive Data Mining.}
The System Provider and/or the Local Authority secretly extract more private information from the collected data, than the Citizen agreed to.
But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties.
This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection.
Or for a Local Authority, who wants to analyze the data collection regarding fare evasion.
However, the Citizen has not agreed to such data processing nor could he, since it is conducted secretly.
This disables a Citizen to control his data adequately.
Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Local Authority}
The threat can be provoked by either lax data handling policies of both System Providers and Local Authorities, or a weak law enforcement of existing supervision.
But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.

\textbf{T3. Data Theft}
An External infiltrates infrastructure in order to steal personal data and sell it on the black market.
Also the External might be motivated politically and wants to harm the reputation of the System Provider or the Local Authority.
Anyways, this threat would be manifested through a technical attack on either hardware (Packet Capture) or software (buffer overflow, SQL injection).
Such a successful attack could harm the reputation of both System Provider and Local Authority.
The technical competence of System Providers would be proven wrong, therefore they would lose professional reputation.
Local Authorities would lose their political reputation, as it would seem like they had endangered data of Citizens, which lose complete control.
Thus this threat is defined by three conflicts: \textit{Privacy vs. Financial Profit}, \textit{Reputation of Service Provider vs. Political Activism} and \textit{Reputation of Local Authority vs. Political Activism}.
Also this threat describes the classical scenario, where attacks are provoked by \textit{Insecure Infrastructure} (SQL injection) and \textit{Insecure Communication} (Packet Capture).

\textbf{T4. Surveillance}

An External infiltrates infrastructure in order to obtain information
about the citizen and exploit it directly.  In this scenario the
external is supposed to have some direct relationship to the citizen
which motivates his interest to obtain personal information.  Examples
could be a public institution that wants to gain information about
planned activities of the citizens (e.g. Nixon's Watergate scandal or
the recent prosecution of Guardian journalists by GHCQ).  Another
example is an insurance company that seeks to get information about
the citizens life-style in relation to the insured risk, like car
accidents or health hazards.

@@

\textbf{T5. Information Leak}

Like an external person the Data Theft Scenario an employee of the
service provider or the Local Authority has selfish interests to gain
money, make political statements or harm his employer.  In order to
pursue this interest he can steal personal data and sell it or release
it to the public.  The corresponding conflicts of interests are:
\textit{Privacy vs. Financial Profit} of the Employee, \textit{Reputation of Service
  Provider vs. Political Activism} of the Employee and \textit{Reputation of Local
  Authority vs. Political Activism} of the Employee.  The vulnerability constitutes of
the existence of \textit{Unhappy employees} itself and possibly \textit{lax access
rules} that enable the employee to obtain large amounts of data
unnoticed.

\textbf{T6. Social Engineering}

This scenario an external manipulates an employee of a Service
Provider or the Local Authority to leak information to the external
person.  It is thus combination of the Data Theft and Information Leak
scenario.
The conflicts of interest are \textit{Privacy vs. Financial Profit} of
the External, \textit{Reputation of Service Provider vs. Political
  Activism} of the external and \textit{Reputation of Local Authority
  vs. Political Activism} of the external.  The exploited
vulnerabilities are, again, the existence of \textit{Unhappy
  employees} and possibly \textit{lax access rules} that enable the
employee to obtain large amounts of data unnoticed.

@@

For the quantification of the loss in case of a threat scenario we use the following rough calibration:
\begin{itemize}
\item 3: High. Leak of information to peers (e.g. public) which impacts citizen.
\item 2: Medium. Undisclosed processing of personal data or disclosure to third parties that are unrelated to subject.
\item 1: Low. Loss of control over data.
\item 0: None
\end{itemize}

For the quantification of likeliness the following scale is used:
\begin{itemize}
\item 4: Always.
\item 3: High. Occurs once in 10 cases
\item 2: Medium. Occurs once in 100 cases
\item 1: Low. Occurs once in 1 million cases
\item 0: Impossible
\end{itemize}

@@

The likeliness, loss and the resulting risks assigned to the threats
are discussed in the following paragraphs and summarized in Figure
\ref{fig:risks}.

\textbf{T1. Insufficient Control Features.}  The occurrence of this
threat is dependent on the design on the system and given in our case,
since we do not give the citizen control over his data once it is
recorded. Therefore the Likeliness is evaluated as $4$ (Always).  The
associated, risk is $1$ Low on our scale, since no direct harm is done to
the citizen by exploiting the data.

Hence the resulting risk is calculated as $4+1 = 5$.

\textbf{T2. Excessive Data Mining.}
We assess the likeliness of excessive data mining to be $3$ High, since
these kind of analysis can be performed within the walls of the service
provider, without somebody else noticing, and the service provider himself
has an interest in this activity.

The associated loss, on the other hand can be substantial (Medium
$2$).  For example when GPS data is linked to data from telephone
books the identity of the citizen can be revealed and personal details
like visits to doctors. In the threat scenario, this information is
not leaked to third parties, (which would justify an even higher loss
assessment), but the very existence of this information violates the
citizens privacy.

Hence the resulting risk is calculated as $3+2 = 5$.

\textbf{T3. Data Theft.}
The likeliness of a targeted attack by a third party is dependent on
the popularity of the offered service and financial value of the
captured information. Moreover, the amount of manual work required to
infiltrate a custom build system is significantly higher that that of
compromising a standard software solution. In the scenario we assume a
moderate popularity in a single metropolitan area, with around 10.000
users and storage of data of only limited financial value (no
addresses, no payment information). Therefore the likeliness
assessment is $1-2$ (Low-Medium).

The harm of leaked information to a criminal party is $3$ High.
Hence the resulting risk is calculated as $4-5$.

\textbf{T4. Surveillance.}  In the surveillance scenario an party
related to the citizen, like a company where he is customer of, or a
government agency, seeks to obtain sensitive information from our
service.

The likeliness of such an intrusion is hard to asses, and depends
again on the popularity of the service. If a high popularity is
reached we have recently learned that spying by government agencies is
very likely to occur. The barrier for companies that do not operate
the infrastructure used to transmit the data a surveillance attack is
however very hard to perform. Therefore we assess the likeliness of
the threat with $1-2$ (Low-Medium).

The harm of leaked information to a related party is $3$ High.
Hence the resulting risk is calculated as $4-5$.

\textbf{T5/6. Information Leak and Social Engineering.}

In our scenario we assume that the culture and ethics inside the
service provider company and local authority are very high, so that
the information leak scenario has a likeliness of $1$ (Low).

The harm of such an information leaked is $3$ High, so that the
resulting risk is calculated as $4$.

\begin{figure}
\centering
\begin{tabular}{|l|l|l|l|l|}
\hline
\textbf{Threat}                   & \textbf{Likeliness} & \textbf{Loss} & \textbf{Risk} & \textbf{Recommendation}
\\\hline
T1. Insufficient Control Features & 4                   & 1             & 5             & R1, R2
\\\hline
T2. Excessive Data Mining         & 3                   & 2             & 5             & R3, R4, R5
\\\hline
T3. Data Theft                    & 1 - 2               & 3             & 4 - 5         & R6
\\\hline
T4. Surveillance                  & 1 - 2               & 3             & 4 - 5         & R7
\\\hline
T5. Information Leak              & 1                   & 3             & 4             & R8
\\\hline
T6. Social Engineering            & 1                   & 3             & 4             & R8
\\\hline
\end{tabular}
\caption{Live+Gov Risk Evaluation and Recommendations}
\label{fig:risks}
\end{figure}

\subsubsection{Privacy Recommendations}
\label{sec:prec}

In the preceding section we have identified the main risks for the
users privacy.  In this section we derive recommendations or
requirements for a system that addresses these risks.  Some of these
requirements are implemented as security measures in our systems and
discussed in the following chapter \ref{chap:impl}.

In order to address the threat with the highers risk, Insufficient
Control (T1) of the citizen, we need to give the citizen back the
control over its data inside the system. The most direct way to do
this is to provide a web-based \emph{Privacy Dashboard (R1)} which allows
the citizen to view, edit and delete all information about his person
that is stored inside the system. Also control applied processing and
disclosure of the data to third parties should be given to the user,
at least in the form of an opt-out or veto option.

A necessary pre-requirement for effective control of the citizen over
his data is information and comprehension of the intended data
capturing and processing steps. Therefore a \emph{Privacy Policy (R2)} that
is easily readable and contains all important information is
essential.  Moreover, the existence of a Privacy Policy is a legal
requirement (cf. Section \ref{EUDIR}).

The threat with the second largest risk is (T2) Excessive Data Mining.
Contrary to common belief, it is neither legal nor ethical to process
personal data for by new methods or for new purposes that were not
stated and explained to the citizen at the time of data
collection. Also the common practice of obtaining far-reaching
permissions from the citizens inside the privacy policy is neither an
ethical or legal solution to the problem (cf. Art. 6 in Section
\ref{EUDIR}).

To address this threat awareness about the limitations of data
processors inside the company is a key element. As one mean to
establish such a culture of privacy respect, we recommend to prepare
an document called \emph{Data Handling Guidelines (R3)} intended for
internal use that explains the concrete processing steps and purposes
that are permitted by the citizens. This guideline should also be
structured in a way that it covers the legal notification requirement
from the EU Data Protection Directive (cf. Section \ref{EUDIR}). In
particular the following information should be provided for each
processing task: The name of the controller, purpose of processing,
description of the data categories, recipients of the data if
disclosed, transfer to third countries and a description of security
of processing.

If further processing should be performed, it is necessary to seek
additional permissions from the citizen. A simple email explaining the
planned processing steps, and \emph{asking for permission (R4)} would be
enough for this purpose. The permission can be given via an embedded
link that shall be followed in order to signal agreement.

An alternative measure to address the risk of excessive data mining is
the \emph{anonymization (R5)} of data. When all direct- or indirect
links to the identity of the person are removed, no violation of the
citizens privacy caused by arbitrary processing. However removing all
such links is a challenging tasks, and full anonymity is often not
achieved, cf. \cite{krumm2009}.

The protection from threat scenario (T3) Data Theft is a case of
classical \emph{IT infrastructure security (R6)}. The storage and processing infrastructure has
to be secured using firewalls, up-to data software versions and proper
authentication mechanisms.

The protection from threat scenario (T4) Surveillance focuses on the
communication channels. They are target of wiretapping attacks by
intermediaries or externals with access to the communication
infrastructure. Strong \emph{encryption (R7)} should be used to make
it harder for externals to read the content of the transmitted data.

Threat scenarios (T5) Information Leak and (T6) Social Engineering
target the vulnerability of unhappy employees. Therefore a trustful,
\emph{healthy company culture (R8)} should be maintained.

In summary we recommend the following measures to secure the citizens
privacy:
\begin{enumerate}
\item[R1] Privacy Dashboard. A tool which allows the citizen to view, edit
  and delete all data personal data that is stored in the system.
\item[R2] Privacy Policy. A document, that informs the citizen about the
  collection and processing of personal information. It should at
  least contain the legally required information.
\item[R3] Issue Data Handling Guidelines that explain the permitted
  processing methods and purposes.
\item[R4] Ask the citizens for permissions before applying further
  processing via Email.
\item[R5] Anonymize personal data before processing.
\item[R6] Securing of Storage and Processing infrastructure using e.g. firewalls.
\item[R7] Securing communication channels using encryption.
\item[R8] Maintain a healthy, trustful relationship with your employees.
\end{enumerate}
The mapping of these recommendations to the threats is summarized in Figure \ref{fig:risks}.
